---
title: "COVID.19 data analysis exercise"
author: "Thomas Fulton"
date: "08/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("utils")
library("ggplot2")
library("tidyverse")
library("lubridate")
```

## R Markdown
```{r}
data <- read.csv("https://opendata.ecdc.europa.eu/covid19/casedistribution/csv", na.strings = "", fileEncoding = "UTF-8-BOM")
```

Download testing data (by week and country)
#```{r}
testing_data <- read.csv("https://opendata.ecdc.europa.eu/covid19/testing/csv", na.strings = "", fileEncoding = "UTF-8-BOM")

#```

Download age data
#```{r}
age_data <- read.csv("https://opendata.ecdc.europa.eu/covid19/agecasesnational/csv", na.strings = "", fileEncoding = "UTF-8-BOM")
#```

Download country response measures
```{r}
prevention_measures <- read.csv("https://www.ecdc.europa.eu/sites/default/files/documents/response_graphs_data_2021-04-15.csv", na.strings = "NA", fileEncoding = "UTF-8-BOM")

prevention_measures$date_end[is.na(prevention_measures$date_end)] <- "2020-12-14"

```


Split the dataset into countries and format date to get just UK data and UK prevention measures
```{r, echo=FALSE}
#summary(data)

data$day <- formatC(data$day, width=2, flag="0")
data$month <- formatC(data$month, width=2, flag="0")
data$formatted_date <- paste(data$year, data$month, data$day, sep = "-")
head(data$formatted_date)


GeoID_dfs <- split(data, data$geoId)
UK_data <- data.frame(GeoID_dfs$UK)


prevention_measures_by_country <- split(prevention_measures, prevention_measures$Country)
UK_measures <- prevention_measures_by_country$`United Kingdom`

```

Add column with the Cumulative_number_for_14_days_of_COVID.19_cases_per_100000 from three weeks earlier, the number of people who could have spread it at -21 days affects the current number of infected people. The lag is seen 3 weeks later (Roy and Ghosh, 2020. https://doi.org/10.1371/journal.pone.0241165). EXPLAIN BETTERR
```{r}

three_weeks_previous_Cumulative_cases <- c(UK_data$Cumulative_number_for_14_days_of_COVID.19_cases_per_100000[22:nrow(UK_data)], rep(0,21))
UK_data$three_weeks_previous_Cumulative_cases <- three_weeks_previous_Cumulative_cases
#head(UK_data$three_weeks_previous_Cumulative_cases)

#colnames(UK_data)
```

Combine UK data and UK prevention measures
```{r}
# get list of dates for each prevention measure in UK_measures
#UK_measures_dates <- list(as.list(1:nrow(UK_measures)))

#nos <- c(1:nrow(UK_measures))
#for (i in seq_along(nos)){
#UK_measures_dates[[i]] = as.list(seq(from = as.Date(UK_measures[i,'date_start']), by = "days", to = as.Date(UK_measures[i,'date_end'])))
#}



#Measures_names <- list(UK_measures$Response_measure)

for (i in c(1:nrow(UK_measures))) {
  dates <- as.data.frame(seq(from = as.Date(UK_measures[i,'date_start']), by = "days", to = as.Date(UK_measures[i,'date_end'])))
  colnames(dates) <- "d"
  
  for (r in c(1:nrow(UK_data))){
    if (UK_data$formatted_date[r] >= min(dates$d) & (UK_data$formatted_date[r] <= max(dates$d))) { 
      UK_data[r,14+i] <- 1
    } 
    else {
      UK_data[r,14+i] <- 0
    }
      
  }
    colnames(UK_data)[14+i] = as.character(UK_measures[i,2])
}

colnames(UK_data) <- make.unique(colnames(UK_data), sep = "_")

```

Combine data from measures that were enforce twice eg. GymsSportsCentres closed from 2020-03-20
to 2020-05-12, and from 2020-11-05 to 2020-12-01. This is currently in two columns.
```{r}

measures_split_into_two_columns <- c(7,11,14,24,29,31,33,38,41)  # check. DONE

# Add presence ("1") of measure for the dates from second column to present in first column
for (i in measures_split_into_two_columns) {
  print(i)
  for (r in c(1:nrow(UK_data))) {
    if (UK_data[r,14+i+1] == 1) {
      UK_data[r,14+i] <- 1
    }
  }
}
#& UK_data$formatted_date[r] < max(dates$d))
```



```{r}
ggplot(data = UK_data, aes(x = as.Date(UK_data$formatted_date), scale_x_date(breaks = waiver()), y = UK_data$Cumulative_number_for_14_days_of_COVID.19_cases_per_100000)) +
  geom_point() 
#str(g)

```



```{r}
ggplot(data = data, aes(x = as.Date(formatted_date), scale_x_date(date_labels = "%b", date_breaks = "1 month"), y = Cumulative_number_for_14_days_of_COVID.19_cases_per_100000, col = countriesAndTerritories)) + 
  geom_line() +
  theme(legend.position = "none")
```


Create linear model
```{r}
lm_model <- lm(Cumulative_number_for_14_days_of_COVID.19_cases_per_100000 ~ three_weeks_previous_Cumulative_cases + BanOnAllEvents + ClosDaycare + ClosDaycarePartial + ClosPrim + ClosPrimPartial + ClosPubAny + ClosPubAnyPartial + ClosSec + ClosSecPartial + EntertainmentVenues + EntertainmentVenuesPartial + GymsSportsCentres + GymsSportsCentresPartial + HotelsOtherAccommodation + HotelsOtherAccommodationPartial + IndoorOver50 + MasksMandatoryClosedSpaces + MasksVoluntaryClosedSpaces + MassGather50 + MassGatherAll + NonEssentialShops + NonEssentialShopsPartial + OutdoorOver50 + PlaceOfWorship + PrivateGatheringRestrictions + RegionalStayHomeOrder + RestaurantsCafes + RestaurantsCafesPartial + SocialCircle + StayHomeGen + StayHomeOrder + StayHomeOrderPartial + StayHomeRiskG + StayHomeRiskGPartial + Teleworking, data = UK_data)

AIC(lm_model)
BIC(lm_model)
anova(lm_model)

lm_model_no_prev3weeks <- lm(Cumulative_number_for_14_days_of_COVID.19_cases_per_100000 ~ BanOnAllEvents + ClosDaycare + ClosDaycarePartial + ClosPrim + ClosPrimPartial + ClosPubAny + ClosPubAnyPartial + ClosSec + ClosSecPartial + EntertainmentVenues + EntertainmentVenuesPartial + GymsSportsCentres + GymsSportsCentresPartial + HotelsOtherAccommodation + HotelsOtherAccommodationPartial + IndoorOver50 + MasksMandatoryClosedSpaces + MasksVoluntaryClosedSpaces + MassGather50 + MassGatherAll + NonEssentialShops + NonEssentialShopsPartial + OutdoorOver50 + PlaceOfWorship + PrivateGatheringRestrictions + RegionalStayHomeOrder + RestaurantsCafes + RestaurantsCafesPartial + SocialCircle + StayHomeGen + StayHomeOrder + StayHomeOrderPartial + StayHomeRiskG + StayHomeRiskGPartial + Teleworking, data = UK_data)

AIC(lm_model_no_prev3weeks)
BIC(lm_model_no_prev3weeks)
anova(lm_model_no_prev3weeks)


```

Test models
```{r}
set.seed(100) 
trainingRowIndex <- sample(1:nrow(UK_data), 0.8*nrow(UK_data))
trainingData <- UK_data[trainingRowIndex, ]
testData  <- UK_data[-trainingRowIndex, ]

lmMod <- lm(Cumulative_number_for_14_days_of_COVID.19_cases_per_100000 ~ BanOnAllEvents + ClosDaycare + ClosDaycarePartial + ClosPrim + ClosPrimPartial + ClosPubAny + ClosPubAnyPartial + ClosSec + ClosSecPartial + EntertainmentVenues + EntertainmentVenuesPartial + GymsSportsCentres + GymsSportsCentresPartial + HotelsOtherAccommodation + HotelsOtherAccommodationPartial + IndoorOver50 + MasksMandatoryClosedSpaces + MasksVoluntaryClosedSpaces + MassGather50 + MassGatherAll + NonEssentialShops + NonEssentialShopsPartial + OutdoorOver50 + PlaceOfWorship + PrivateGatheringRestrictions + RegionalStayHomeOrder + RestaurantsCafes + RestaurantsCafesPartial + SocialCircle + StayHomeGen + StayHomeOrder + StayHomeOrderPartial + StayHomeRiskG + StayHomeRiskGPartial + Teleworking, data=trainingData)

Cumulative_cases_Pred <- predict(lmMod, testData)
summary(lmMod)

actuals_preds <- data.frame(cbind(actuals=testData$Cumulative_number_for_14_days_of_COVID.19_cases_per_100000, predicteds=Cumulative_cases_Pred))
actuals_preds[is.na(actuals_preds)] <- 0  # one NA which should have been 0 in 

correlation_accuracy <- cor(actuals_preds)

min_max_accuracy <- mean(apply(actuals_preds, 1, min) / apply(actuals_preds, 1, max))
mape <- mean(abs((actuals_preds$predicteds - actuals_preds$actuals))/actuals_preds$actuals)  


```


Check assumptions
```{r}

mean(lm_model_no_prev3weeks$residuals)  # mean of residuals is roughly equal to 0

plot(lm_model_no_prev3weeks)

acf(lm_model_no_prev3weeks$residuals)

```

Refinement:
Model selection and testing
```{r, echo=FALSE}
selectedMod <- step(lmMod)

Cumulative_cases_Pred <- predict(selectedMod, testData)
summary(selectedMod)

actuals_preds <- data.frame(cbind(actuals=testData$Cumulative_number_for_14_days_of_COVID.19_cases_per_100000, predicteds=Cumulative_cases_Pred))
actuals_preds[is.na(actuals_preds)] <- 0  # one NA which should have been 0 in 

correlation_accuracy <- cor(actuals_preds)

min_max_accuracy <- mean(apply(actuals_preds, 1, min) / apply(actuals_preds, 1, max))
mape <- mean(abs((actuals_preds$predicteds - actuals_preds$actuals))/actuals_preds$actuals)

```

Create generalized linear model?

Include the effect of cumulative no. of cases from 3 weeks ago: no. cumulative cases from 3 weeks earlier as a variable for the lag in effect of prevention measures.


Find the discriminatory variables - ie. which prevention measures were enacted at the same time

Plot the dates on top of cumulative cases line graph: shade area under what date each group of prevention measures was implemented.


WOE - feature selection / VIF? 

AMOVA /perMANOVA

Linear regression categorical

test variables against no. cases, then maybe no. deaths?


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.





