---
title: "transformed_COVID_data_analysis"
author: "Thomas Fulton"
date: "27/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```






### Create linear model
```{r, results="hide" }
#predictors <- list()
#for (i in colnames(UK_data)) {
#  predictors <- c(predictors, " + ", i)
#}

#predictors <- paste(unlist(predictors), collapse = "")  # get string of predictors for lm function

lm_model <- lm(root_cumul_cases ~root_prev_cumul_cases + BanOnAllEvents + ClosDaycare + ClosDaycarePartial + ClosPubAny + ClosPubAnyPartial + ClosSec + ClosSecPartial + EntertainmentVenues + EntertainmentVenuesPartial + GymsSportsCentres + GymsSportsCentresPartial + HotelsOtherAccommodation + HotelsOtherAccommodationPartial + IndoorOver50 + MasksMandatoryClosedSpaces + MasksVoluntaryClosedSpaces + MassGather50 + NonEssentialShops + NonEssentialShopsPartial + OutdoorOver50 + PlaceOfWorship + PrivateGatheringRestrictions + RegionalStayHomeOrder + RestaurantsCafesPartial + SocialCircle + StayHomeGen + StayHomeOrder + StayHomeOrderPartial + StayHomeRiskG + StayHomeRiskGPartial, data = UK_data)
summary(lm_model)
AIC(lm_model)
BIC(lm_model)
anova(lm_model)

```

```{r}
plot(lm_model)

```

Create Model exluding_cumulative_previous_3_weeks variable
```{r}

lm_model_no_prev3weeks <- lm(root_cumul_cases ~ BanOnAllEvents + ClosDaycare + ClosDaycarePartial + ClosPubAny + ClosPubAnyPartial + ClosSec + ClosSecPartial + EntertainmentVenues + EntertainmentVenuesPartial + GymsSportsCentres + GymsSportsCentresPartial + HotelsOtherAccommodation + HotelsOtherAccommodationPartial + IndoorOver50 + MasksMandatoryClosedSpaces + MasksVoluntaryClosedSpaces + MassGather50 + NonEssentialShops + NonEssentialShopsPartial + OutdoorOver50 + PlaceOfWorship + PrivateGatheringRestrictions + RegionalStayHomeOrder + RestaurantsCafesPartial + SocialCircle + StayHomeGen + StayHomeOrder + StayHomeOrderPartial + StayHomeRiskG + StayHomeRiskGPartial, data = UK_data)

summary(lm_model_no_prev3weeks)
AIC(lm_model_no_prev3weeks)
BIC(lm_model_no_prev3weeks)
anova(lm_model_no_prev3weeks)


```

Test models
```{r}
set.seed(100) 
trainingRowIndex <- sample(1:nrow(UK_data), 0.8*nrow(UK_data))
trainingData <- UK_data[trainingRowIndex, ]
testData  <- UK_data[-trainingRowIndex, ]

Cumulative_cases_Pred <- predict(lm_model, testData)
#summary(lm_model_no_prev3weeks)

actuals_preds <- data.frame(cbind(actuals=testData$root_cumul_cases, predicteds=Cumulative_cases_Pred))
actuals_preds[is.na(actuals_preds)] <- 0  # one NA which should have been 0 in 

correlation_accuracy <- cor(actuals_preds)

min_max_accuracy <- mean(apply(actuals_preds, 1, min) / apply(actuals_preds, 1, max))
mape <- mean(abs((actuals_preds$predicteds - actuals_preds$actuals))/actuals_preds$actuals)  


```


##Check assumptions

Residuals are roughly normally distributed
Check mean of residuals is roughly equal to 0
- mean of residuals is roughly equal to 0
```{r}
# 
mean(lm_model_no_prev3weeks$residuals)  # mean of residuals is roughly equal to 0
# [1] 1.634789e-15
```

Check Homoscedasticity (random noise) of residuals or equal variance
- Residuals vs fitted is approximately flat: homoscedastic, but standardised residuals vs fitted values show slight heteroscedasticicity.
```{r}
# Residuals are the difference between the actual observed response values and the response values that the model predicted.
#par(mfrow=c(2,2))
plot(lm_model)

```

Check for autocorrelation - should hopefully be accounted for by root_prev_cumul_cases variable.

```{r}

acf(lm_model_no_prev3weeks$residuals)
#runs.test(lm_model2$residuals)

```

Normality of Residues
- Loosely normally distributed (plot3)
```{r}
plot(lm_model)
```



## Refinement of model:
1. Step wise regression to select factors for a model with lower AIC
2. Adjust for Multicoliniarity
```{r, echo=FALSE, results="hide"}
selectedMod <- step(lm_model)
anova(selectedMod)  # all
summary(selectedMod)

```
step wise regression reduced number of factors to 11 factors. 10 out of 11 are significantly influencial (p<=0.5).  

Test for multicoliniarity (Code from http://r-statistics.co/Model-Selection-in-R.html)

```{r}
#step wise regression reduced number of factors to 11 factors. 10 out of 11 are significantly influencial (p<=0.5).  
#Unfortuanatey did not have time to test:/

#Test for multicoliniarity (Code from http://r-statistics.co/Model-Selection-in-R.html)

all_vifs <- car::vif(selectedMod)

signif_all <- names(all_vifs)
while(any(all_vifs > 4)){
  var_with_max_vif <- names(which(all_vifs == max(all_vifs)))  # get the var with max vif
  signif_all <- signif_all[!(signif_all) %in% var_with_max_vif]  # remove
  myForm <- as.formula(paste("root_cumul_cases ~ ", paste (signif_all, collapse=" + "), sep=""))  # new formula
  selectedMod <- lm(myForm, data=UK_data)  # re-build model with new formula
  all_vifs <- car::vif(selectedMod)
}
summary(selectedMod)
Anova_selected_Mod <- anova(selectedMod)
Anova_selected_Mod
write.csv(Anova_selected_Mod, "~/Downloads/Anova_selected_Mod.csv")
```

## Assess improved model
Test models
```{r}
set.seed(100) 
trainingRowIndex <- sample(1:nrow(UK_data), 0.8*nrow(UK_data))
trainingData <- UK_data[trainingRowIndex, ]
testData  <- UK_data[-trainingRowIndex, ]

sel_Cumulative_cases_Pred <- predict(selectedMod, testData)
summary(selectedMod)

sel_actuals_preds <- data.frame(cbind(actuals=testData$root_cumul_cases, predicteds=sel_Cumulative_cases_Pred))
sel_actuals_preds[is.na(actuals_preds)] <- 0  # one NA which should have been 0 in 

sel_correlation_accuracy <- cor(actuals_preds)
sel_correlation_accuracy
sel_min_max_accuracy <- mean(apply(actuals_preds, 1, min) / apply(actuals_preds, 1, max))
sel_mape <- mean(abs((actuals_preds$predicteds - actuals_preds$actuals))/actuals_preds$actuals)  


```


#Check assumptions for improved model

Residuals are roughly normally distributed
Check mean of residuals is roughly equal to 0
- mean of residuals is roughly equal to 0
```{r}
# 
mean(selectedMod$residuals)  # mean of residuals is roughly equal to 0
# [1] 1.634789e-15
```

Check Homoscedasticity (random noise) of residuals or equal variance
- Residuals vs fitted is approximately flat: homoscedastic, but standardised residuals vs fitted values show slight heteroscedasticicity.
```{r}
# Residuals are the difference between the actual observed response values and the response values that the model predicted.
#par(mfrow=c(2,2))
plot(selectedMod)
summary(selectedMod)
```

Check for autocorrelation - should hopefully be accounted for by root_prev_cumul_cases variable.
Still autocorrelation - for future could try and improve by introducing a better lag variable by making multiple models with different lags eg. from 1 week-4 weeks and comparing models.
```{r}

acf(selectedMod$residuals)
#runs.test(lm_model2$residuals)

```

Normality of Residues
- Not very normally distributed. 
```{r}
plot(lm_model)
summary(selectedMod)
AIC(selectedMod)
```


MARS method for variable importance based on Generalized cross validation (GCV), number of subset models the variable occurs (nsubsets) and residual sum of squares (RSS).
```{r}
regressor <- earth(root_cumul_cases ~ root_prev_cumul_cases, ClosDaycarePartial, ClosPubAnyPartial, EntertainmentVenuesPartial, MasksVoluntaryClosedSpaces, NonEssentialShops, PlaceOfWorship, RegionalStayHomeOrder, SocialCircle, data=UK_data) # build model

ev <- evimp(selectedMod) # estimate variable importance

plot (ev)




```



